---
title: "schedule"
format: 
  html: 
    page-layout: full
---

```{r load-libraries, echo = F, message = F, eval = F}
library(tidyverse)
library(knitr)
library(emo)
```

```{r smart-date-setup, echo = F, message = F, eval = F}
# set these:
firstDay = as.Date("2023-01-12")
firstDayID = "Thu"
lastDay = as.Date("2023-04-26")
lectureDays = c("Wed", "Fri")
labDays = c("Tue")
holidays = as.Date(c("2023-01-16", "2023-03-14", "2023-03-15", "2023-03-17"))

### render table
allDates = seq(from = firstDay, to = lastDay, by = 'day')
schedule = data.frame(allDates,
                        dayOfWeek = weekdays(allDates, abbreviate=T)) %>%
  filter(dayOfWeek %in% c(lectureDays, labDays))
schedule = schedule %>%
  mutate(lab = ifelse(dayOfWeek %in% labDays, TRUE, FALSE)) %>%
  mutate(Topic = ifelse(allDates %in% holidays, "NO CLASS", "")) %>%
  mutate(Date = format(allDates, format = "%b %d")) %>%
  mutate(Date = paste(dayOfWeek, Date)) %>%
  mutate(Week = rep("", nrow(schedule))) %>%
  mutate(Prepare = rep("", nrow(schedule))) %>%
  mutate(AE = rep("", nrow(schedule))) %>%
  mutate(Assignment = rep("", nrow(schedule))) %>%
  mutate(Exam = rep("", nrow(schedule))) %>%
  mutate(Project = rep("", nrow(schedule))) %>%
  select(Week, Date, Topic, Prepare, AE, Assignment, Exam, Project)
# to be pasted into console then pasted into markdown:
schedule %>%
  kable(format = "markdown")
```

| Week | Date       | Topic                                     |              Notes              | Lab                      | Exam | Project |
|:---------|:------------|:-------------|:--------:|:------------|:--------------------|:---------|
| 0    | Fri Jan 13 | welcome                                   |  [💻](/slides/01-welcome.html)  |                          |      |         |
| 1    | Tue Jan 17 | git                                       | [💻](/slides/lab0-welcome.html) | [lab 0](/labs/lab0.html) |      |         |
|      | Wed Jan 18 | control flow, loops and functions         |    [💻](/slides/lec02.html)     |                          |      |         |
|      | Fri Jan 20 | NAs, lists, NULL and attributes           |    [💻](/slides/lec03.html)     |                          |      |         |
| 2    | Tue Jan 24 | if it's a function, fix it                |                                 | [lab 1](/labs/lab1.html) |      |         |
|      | Wed Jan 25 | Object oriented programming               |    [💻](/slides/lec04.html)     |                          |      |         |
|      | Fri Jan 27 | tidy data and dplyr           |  [💻](/slides/lec05.html)                                |                          |      |         |
| 3    | Tue Jan 31 | computational masonry                                          |                                 |   [lab 2](/labs/lab2.html)                       |      |         |
|      | Wed Feb 01 | subsetting                                   | [💻](/slides/lec06.html)                                |                          |      |         |
|      | Fri Feb 03 | ggplot2          | [💻](/slides/lec07.html)                                |                          |      |         |
| 4    | Tue Feb 07 | dream of plots                                          |                                 |  [lab 3](/labs/lab3.html)                        |      |         |
|      | Wed Feb 08 | functional programming and purrr   |  [💻](/slides/lec08.html)                               |                          |      |         |
|      | Fri Feb 10 | regular expressions; UNIX toolkit                     |                                 |                          |      |         |
| 5    | Tue Feb 14 |                                           |                                 |                          |      |         |
|      | Wed Feb 15 | complexity, profiling and parallelization |                                 |                          |      |         |
|      | Fri Feb 17 | Review                                    |                                 |                          | exam 1 release  |         |
| 6    | Tue Feb 21 | no lab: exam 1                                        |                                 | exam 1 due                         |      |         |
|      | Wed Feb 22 | statistical modeling overview             |                                 |                          |      |         |
|      | Fri Feb 24 | intro to optimization                     |                                 |                          |      |         |
| 7    | Tue Feb 28 |                                           |                                 |                          |      |         |
|      | Wed Mar 01 | univariate Newton-Raphson                 |                                 |                          |      |         |
|      | Fri Mar 03 | multivariate Newton-Raphson examples      |                                 |                          |      |         |
| 8    | Tue Mar 07 |                                           |                                 |                          |      |         |
|      | Wed Mar 08 | MM algorithm I                            |                                 |                          |      |         |
|      | Fri Mar 10 | MM algorithm II                           |                                 |                          |      |         |
| 9    | Tue Mar 14 | NO CLASS: Spring break                    |                                 |                          |      |         |
|      | Wed Mar 15 | NO CLASS: Spring break                    |                                 |                          |      |         |
|      | Fri Mar 17 | NO CLASS: Spring break                    |                                 |                          |      |         |
| 10   | Tue Mar 21 |                                           |                                 |                          |      |         |
|      | Wed Mar 22 | EM algorithm                              |                                 |                          |      |         |
|      | Fri Mar 24 | matrices and BLAS                         |                                 |                          |      |         |
| 11   | Tue Mar 28 |                                           |                                 |                          |      |         |
|      | Wed Mar 29 | project tips                              |                                 |                          |      |         |
|      | Fri Mar 31 | shiny                                     |                                 |                          |      |         |
| 12   | Tue Apr 04 |                                           |                                 |                          |      |         |
|      | Wed Apr 05 | Monte Carlo integration                   |                                 |                          |      |         |
|      | Fri Apr 07 | importance sampling                       |                                 |                          |      |         |
| 13   | Tue Apr 11 |                                           |                                 |                          |      |         |
|      | Wed Apr 12 | spatial data                              |                                 |                          |      |         |
|      | Fri Apr 14 | covariance                                |                                 |                          |      |         |
| 14   | Tue Apr 18 |                                           |                                 |                          |      |         |
|      | Wed Apr 19 | APIs                                      |                                 |                          |      |         |
|      | Fri Apr 21 | SQL                                       |                                 |                          |      |         |
| 15   | Tue Apr 25 | project presentations                     |                                 |                          |      |         |
|      | Wed Apr 26 | special topic                             |                                 |                          |      |         |
