---
title: "Object oriented programming"
author: "Dr. Alexander Fisher"
date: "January 25, 2023"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

# Review of types

## R's type hierarchy

-   from most to least complex: class $\rightarrow$ mode $\rightarrow$ type

```{r}
#| echo: false
f = function(x) x^2
x = factor("A")
l = list(1, "A")
```

| value             | `typeof()`       | `mode()`       | `class()`       |
|:------------------|:-----------------|:---------------|:----------------|
| `TRUE`            | `r typeof(TRUE)` | `r mode(TRUE)` | `r class(TRUE)` |
| `1`               | `r typeof(1)`    | `r mode(1)`    | `r class(1)`    |
| `1L`              | `r typeof(1L)`   | `r mode(1L)`   | `r class(1L)`   |
| `"A"`             | `r typeof("A")`  | `r mode("A")`  | `r class("A")`  |
| `NULL`            | `r typeof(NULL)` | `r mode(NULL)` | `r class(NULL)` |
| `list(1, "A")`    | `r typeof(l)`    | `r mode(l)`    | `r class(l)`    |
| `factor("A")`     | `r typeof(x)`    | `r mode(x)`    | `r class(x)`    |
| `function(x) x^2` | `r typeof(f)`    | `r mode(f)`    | `r class(f)`    |
| `\+`              | builtin          | function       | function        |
| `\[`              | special          | function       | function        |

. . .

objects have `class` even if no formal `class` attribute (seen with `attributes()`)

# S3 Object System

## What is S3?

- S3 is R's core object oriented implementation

- S3 is the only OO system used in the base and stats packages, and it’s the most commonly used system in CRAN packages.

- S3 is more flexible (read also: easier to break) than traditional object oriented programming you may have seen before (e.g. Java).

. . .

Working model: S3 is a system where we attach metadata to an object via the `class` attribute.

## S3 class specialization

```{r}
x = c("A","B","A","C")
```

. . .

```{r}
print( x )
```

. . .

```{r}
print( factor(x) )
```

. . .

```{r}
print( unclass( factor(x) ) ) # unclass strips the class attribute
```

. . .

```{r}
print.default( factor(x) )
```

. . .


## Other examples

::: columns
::: {.column width="50%"}
```{r}
mean
```
:::

::: {.column width="50%"}
```{r}
t.test
```
:::
:::

. . .

Not all base functions use this approach,

```{r}
sum
```

. . .

::: callout-note
The `...` (dot-dot-dot) argument means the function can take any number of additional arguments. This is called a "variable argument" and it is commonly used so you pass variables to additional nested functions. For example:

```{r}
#| eval: false
f1 =  function(y = 0, z = 0) {
  return(y + z)
}

f2 <- function(x, ...) {
  return(x + f1(...))
}

f2(x = 1, y = 2, z = 3)
```
:::

## Generics (what's going on!?)

- An S3 object behaves differently from its underlying base type whenever it's passed to a **generic** function

The easiest way to tell if a function is a generic is to use `sloop::ftype()` and look for "generic" in the output, e.g.

```{r}
library(sloop)
ftype(print)
```

. . .

Fundamentally, the generic function dispatches to a specialized version of the function depending on the class of its input. `UseMethod()` tells a generic to look for the specialized version.

The syntax for specialized functons is: `<generic>.<class>`




