---
title: "Control flow, loops, and functions"
author: "Dr. Alexander Fisher"
date: "January 18, 2023"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

# Control flow

## `if` statements

```{r}
#| eval: false
if (condition) {
  do stuff # when condition is TRUE
}
```

### Examples

```{r}
x = c(1,2,3)
if (2 %in% x) {
  print("2 is in x!")
}
```

```{r}
if (-6) {
  print("Other types are coerced to logical if possible.")
}
```

```{r}
if (5 %in% x) 
  print("5 is in x!")
```

. . .

if the statement is `FALSE`, {the code} does not execute

## `if` is not vectorized

While many operators and functions in `R` are vectorized,

```{r}
x = c(1,2,3)
```

```{r}
exp(x)
log(x)
x + 2
```

. . .

`if` statements are **not**

. . .

```{r}
if (x == 1) {
  print("x is 1!")
}
```

## Collapsing logical vectors

```{r}
x = c(1, 2, 3)
x > 0
```

`any()` and `all()` can help us collapse this to a single argument (like "or"/"and" logic)

```{r}
any(x > 0)
all(x > 0)
```

. . .

```{r}
if (any(x > 0)) {
  print("At least one element of x is greater than 0.")
}
```

## `else if`, `else` and `ifelse`

::: columns
::: {.column width="33%"}
```{r}
x = 3
if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
:::

::: {.column width="33%"}
```{r}
x = 0
if (x < 0) {
  "x is negative"
} else if (x > 0) {
  "x is positive"
} else {
  "x is zero"
}
```
:::

::: {.column width="33%"}
```{r}
x = -1
ifelse(x > 0,
       "positive",
       "not positive")
```
:::
:::

# Error checking

## `stop` and `stopifnot`

We often need to validate user input and function arguments. If our validation fails, we want to report the error and stop execution.

```{r}
ok = FALSE
```

```{r}
#| error: true
if (!ok)
  stop("Things are not ok.")
```

```{r}
#| error: true
stopifnot(ok)
```

::: callout-note
An error (like the one generated by stop) will prevent a quarto document from rendering unless `#| error: true` is set for that code chunk
:::

## Placing checkpoints

Always place checkpoints upstream (find errors as quickly as possible).

::: columns
::: {.column width="50%"}
Bad checkpoint placement

```{r}
#| eval: false
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
} else if (condition_error) {
  stop("Condition error occured")
}


```
:::

::: {.column width="50%"}
Good checkpoint placement

```{r}
#| eval: false
# Do stuff better
if (condition_error) {
  stop("Condition error occured")
}
if (condition_one) {
  ##
  ## Do stuff
  ##
} else if (condition_two) {
  ##
  ## Do other stuff
  ##
}
```
:::
:::

## Exercise 1

Consider two vectors, x and y, each of length one. Write a set of conditionals that satisfy the following.

-   If x is positive and y is negative or y is positive and x is negative, print "knits".
-   If x divided by y is positive, print "stink".
-   Stop execution if x or y are zero.

Test your code with various x and y values. Where did you place the stop execution code?

# Loops

## Loop types

R supports three types of loops: `for`, `while`, and `repeat`.

```{r}
#| eval: false
for (item in vector) {
##
## Iterate this code
##
}
```

```{r}
#| eval: false
while (we_have_a_true_condition) {
##
## Iterate this code
##
}
```

```{r}
#| eval: false
repeat {
##
## Iterate this code
##
}
```

## for loops

`for` loops are used to iterate over items in a vector. They have the following basic form:

```{r}
#| eval: false
for (item in vector) perform_action
```

```{r}
for (nakama in c("Luffy", "Nami", "Zoro")) {
  print(nakama)
}
```

```{r}
for (i in 1:4) {
  log(i)
}
```

. . .

Automatic printing is turned off in loops.

## while loops

`while` loops interate until a condition is false

```{r}
squares = rep(0, 5)
squares
```

```{r}
i = 1
while (i < 6) {
  squares[i] = i^2
  i = i + 1
}
squares
```

## repeat loops

`repeat` loops repeatedly iterate code until a `break` is reached.

```{r}
i = 1
squares = rep(0, 5)
repeat {
squares[i] = i ^ 2
i = i + 1
if (i > 5) {break}
}
squares
```

## loop keywords: `next` and `break`

-   `next` exits the current iteration and advances the looping index
-   `break` exits the loop
-   both `break and`next apply only to the innermost of nested loops.

```{r}
for (i in 1:10) {
  if (i %% 2 == 0) {next}
  print(paste("Number", i, "is odd."))
  if (i %% 7 == 0) {break}
  }
```

## Auxiliary loop functions

You may want to loop over indices of an object as opposed to the object's values. To do this, consider using one of `length()`, `seq()`, `seq_along()`, and `seq_len()`.

::: columns
::: {.column width="50%"}
::: columns
`seq_along(x)` is preferred to `1:length(x)` e.g.

```{r}
x = list()
length(x)
1:length(x)
seq_along(x)
```
:::
:::

::: {.column width="50%"}
Many ways to generate sequences...

```{r}
1:5
seq(1:5)
seq_len(5)

```
:::
:::

## Exercises

### Exercise 2

Consider the vector x below.

```{r}
x = c(3, 4, 12, 19, 23, 49, 100, 63, 70)
```

Write R code that prints the perfect squares in x.

### Exercise 3

Consider `z = c(-1, .5, 0, .5, 1)`. Write R code that prints the smallest non-negative integer k satisfying the inequality 

$$
|cos(k)−z|<0.001
$$ 

for each component of z.

# Functions

## Function composition

A function is comprised of arguments (formals), and code (body).

```{r}
quadraticRoots = function(a, b, c) {
  x1 = (-b + sqrt((b^2) - (4*a*c))) / (2*a)
  x2 = (-b - sqrt((b^2) - (4*a*c))) / (2*a)
  return(c(x1, x2))
}

quadraticRoots(1, -2, -3)
```

```{r}
formals(quadraticRoots)
body(quadraticRoots)
```

## Returns

There are two approaches to returning values from functions in R - explicit and implicit returns.

Explicit - using one or more return function calls

```{r}
f = function(x) {
  return(x * x)
}
f(2)
```

. . .

Implicit - return value of the last expression is returned.

```{r}
g = function(x) {
  x * x
}
g(3)
```

## Invisible returns

Many functions in R make use of an invisible return value

-   visible

```{r}
f = function(x) {
  print(x)
}
y = f(1)
y
```

. . .

-   invisible

```{r}
g = function(x) {
  invisible(x)
}
g(2)
z = g(2)
z
```

## Arguments

When defining a function we explicitly define names for the arguments, which become variables within the scope of the function.

When calling a function we can use these names to pass arguments in an alternative order.

```{r}
f = function(x, y, z = 1) { # z defaults to 1
  paste0("x=", x, " y=", y, " z=", z)
}
```

::: columns
::: {.column width="50%"}
```{r}
f(1, 2, 3)
f(z = 1, x = 2, y = 3)
f(y = 2, 1, 3)
```
:::

::: {.column width="50%"}
```{r}
#| error: true
f(1)
f(1, 2)
f(1, 2, 3, 4)
f(1 , 2, m = 3)
```
:::
:::

## Scope

R has generous scoping rules, if it can't find a variable in the current scope (e.g. a function's body) it will look for it in the next higher scope, and so on.

```{r}
y = 1
f = function(x) {
  x + y
}
f(3)
```

. . .

```{r}
y = 1
g = function(x) {
  y = 2
  x + y
}
g(3)
```

. . .

Additionally, variables defined within a scope only persist for the duration of that scope, and do not overwrite variables at a higher scopes

## Lazy evaluation

Arguments to R functions are not evaluated until needed.

```{r}
#| error: true
f = function(a, b, x) {
  print(a)
  print(b ^ 2)
  0 * x
}
f(5, 6)
```

## Function forms

| Form        | Description                  | Example(s)                |
|-------------|------------------------------|---------------------------|
| Prefix      | name comes before arguments  | `log(x, base = exp(1))`   |
| Infix       | name between arguments       | `+, %>%, %/%`             |
| Replacement | replace values by assignment | `names(x) <- c("a", "b")` |
| Special     | all others not defined above | `[[, for, break, (`       |

## Help

To get help on any function, type ?fcn_name in your console, where fcn_name is the function's name. For infix, replacement, and special functions you will need to surround the function with backticks.

```{r}
#| eval: false
?mean
?`for`
?`+`
```

For functions not in the base package, you can generally see their implementation by entering the function name without parentheses (or using the body function).

```{r}
lm |>
  head()
```

## Function best practices

-   Write a function when you have copied code more than twice.

-   Try to use a verb for your function's name.

-   Keep argument names short but descriptive.

-   Add code comments to explain the "why" of your code.

-   Link a family of functions with a common prefix: `pnorm()`, `pbinom()`, `ppois()`.

-   Keep data arguments first, then other required arguments, then followed by default arguments. The ... argument can be placed last.

## A summary of R  

> To understand computations in R, two slogans are helpful:
>
> Everything that exists is an object.
>
> Everything that happens is a function call.
>
> --- John Chambers

John McKinley Chambers is the creator of the S programming language, and core member of the R programming language project. The R programming is often called a successor to the S programming language.