---
title: "Importance sampling"
author: "Dr. Alexander Fisher"
execute:
  warning: true
format: 
    revealjs:
      smaller: true
---

# Announcements

- Reminders:
  - lab 6 due tomorrow
  - exam 2 on Friday

. . . 

- quiz

## Last time

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
```


- We considered computing the integral

$$
V = \int_1^3 x~e^{-\frac{(x-4)^2}{2}} \frac{1}{\sqrt{2\pi}} e^{-\frac{x^2}{2}} dx
$$
Here, $p(x)$ is the standard normal density and 

$$
h(x) = I(x \in (1, 3))~x~e^{-\frac{(x-4)^2}{2}}
$$

. . . 

So we can view this as $\mathbb{E}~h(x) \approx \frac{1}{N} \sum_{i=1}^N h(x)$ where $X$ is standard normal.


## $p(x)$ off from $h(x)$

- Note: here $h(x) = x~e^{-\frac{(x-4)^2}{2}} \frac{1}{\sqrt{2\pi}}$ (i.e. I am dropping the indicator function for illustrative purposes)

```{r}
#| echo: false
#| message: false
#| warning: false
h = function(x) {
  term = - 1 *((x - 4) ^ 2) / 2
  x * exp(term)
}

ggplot() +
 xlim(-4, 8) +
  geom_function(fun = h, aes(col = "h(x)")) +
  geom_function(fun = dnorm, args = list(mean = 0, sd = 1), aes(col = 'p(x)')) +
  labs(x = "x", y = "") +
  theme_minimal() +
  scale_color_manual(values = c("red", "steelblue"))
```
## Samples not good for $h(x)p(x)$

:::panel-tabset

## plot

```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(2)
hp = function(x) {
  h(x) * dnorm(x)
}

N = 50
points = data.frame(x = rnorm(N, 0, 1),
                    y = rep(0, N))

ggplot() +
 xlim(-4, 8) +
  geom_function(fun = hp) +
  geom_point(data = points, aes(x = x, y = y)) +
  labs(x = "x", y = "") +
  theme_minimal() + 
  labs(title = "h(x)p(x)") +
  stat_function(fun = hp, 
                xlim = c(1,3),
                geom = "area",
                fill = "steelblue", alpha = 0.5)
```
## code
```{r}
#| eval: false
set.seed(2)
hp = function(x) {
  h(x) * dnorm(x)
}

N = 50
points = data.frame(x = rnorm(N, 0, 1),
                    y = rep(0, N))

ggplot() +
 xlim(-4, 8) +
  geom_function(fun = hp) +
  geom_point(data = points, aes(x = x, y = y)) +
  labs(x = "x", y = "") +
  theme_minimal() + 
  labs(title = "h(x)p(x)") +
  stat_function(fun = hp, 
                xlim = c(1,3),
                geom = "area",
                fill = "steelblue", alpha = 0.5)
```
:::

## Resulting in large approximation error

::: panel-tabset

## plot
```{r}
#| echo: false
#| message: false
#| warning: false
set.seed(2)

# N = 10000
# x = rnorm(N, 0, 1)
# 
# h1 = function(x) {
# z = h(x)
# z[x < 1] = 0
# z[x > 3] = 0
#   return(z)
# }
# 
# estimate = vector(length = N)
# for (n in 1:N) {
#   estimate[n] = mean(h1(x[1:n]))
# }
# 
# V = data.frame(x = seq(N),
#                y = estimate)

V = read_csv("../data/lectureMonteCarlo_V.csv")

V %>%
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_minimal() +
  labs(x = "N", y = "sample mean") +
  scale_x_continuous(trans='log10')
```

## code
```{r}
#| eval: false
library(magrittr)
set.seed(2)
N = 10000
x = rnorm(N, 0, 1)

h1 = function(x) {
z = h(x)
z[x < 1] = 0
z[x > 3] = 0
  return(z)
}

estimate = vector(length = N)
for (n in 1:N) {
  estimate[n] = mean(h1(x[1:n]))
}

V = data.frame(x = seq(N),
               y = estimate)

V %>%
  ggplot(aes(x = x, y = y)) +
  geom_line() +
  theme_minimal() +
  labs(x = "N", y = "sample mean") +
  scale_x_continuous(trans='log10')
```

:::

## A better strategy

- Notice that

$$
V = \frac{e^{-4}}{\sqrt{2}}
\int_1^3 x \frac{1}{\sqrt{2\pi(1/2)}}
e^{-\frac{(x-2)^2}{2(1/2)}}
$$
- So we can approximate $V$ by

$$
\frac{e^{-4}}{\sqrt{2}}~\frac{1}{N}\sum_{i = 1}^N X_i~I(X_i\in(1,3))
$$

where $X_i \overset{\mathrm{iid}}{\sim} N(2, (1/\sqrt{2})^2))$

## Comparison

::: panel-tabset

## Plot

```{r}
#| echo: false
#| warning: false
set.seed(2)
# 
# N = 10000
# x = rnorm(N, 2, sqrt(1/2))
# 
# hnew = function(x) {
# x[x < 1] = 0
# x[x > 3] = 0
#   return(x)
# }
# 
# C = exp(-4) / sqrt(2)
# 
# estimate = vector(length = N)
# for (n in 1:N) {
#   estimate[n] = mean(hnew(x[1:n]))
# }
# 
# estimate = estimate*C
# 
# V2 = data.frame(x = seq(N),
#                y = estimate)

V2 = read_csv("../data/lectureMonteCarlo_V2.csv")

V %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(color = "Naive approach")) +
  geom_line(data = V2, aes(x = x, y = y, col = "New sampler")) +
  theme_minimal() +
  labs(x = "N", y = "sample mean") +
  scale_x_continuous(trans='log10')
```

## Code
```{r}
#| eval: false
library(magrittr)
set.seed(2)
N = 10000
x = rnorm(N, 2, sqrt(1/2))

hnew = function(x) {
x[x < 1] = 0
x[x > 3] = 0
  return(x)
}

C = exp(-4) / sqrt(2)
# C = sqrt(1/2)

estimate = vector(length = N)
for (n in 1:N) {
  estimate[n] = mean(hnew(x[1:n]))
}

estimate = estimate*C

V2 = data.frame(x = seq(N),
               y = estimate)

V %>%
  ggplot(aes(x = x, y = y)) +
  geom_line(aes(color = "Naive approach")) +
  geom_line(data = V2, aes(x = x, y = y, col = "New sampler")) +
  theme_minimal() +
  labs(x = "N", y = "sample mean") +
  scale_x_continuous(trans='log10')
```
:::

## Importance sampling identity

- One can always write $V$ as

$$
V = \int \frac{h(x)p(x)}{q(x)} q(x) dx
$$
for any density $q$ with $q(x) > 0$ on $\mathcal{X} = \{x : h(x)p(x) \neq 0\}$

. . . 

- Thus we can always approximate $V$ by 

$$
\hat{V}_q = \frac{1}{N} \sum_{i =1}^N w_q(X_i)h(X_i)
$$

where $w_q(x) = p(x) / q(x)$ and $X_i \overset{\mathrm{iid}}{\sim} q$. $w_q(x)$ is called the "importance weight".

. . . 

- The efficiency of $\hat{V}_q$ is inversely proportional to its variance:

$$
\sigma^2_q = \mathbb{V}ar[\hat{V}_q] = \int \frac{h(x)p(x)}{q(x)}h(x)p(x)dx - V^2
$$
If $h > 0$ on $\mathcal{X}$, then $\sigma_q^2$ is minimized at $q(x) \propto h(x)p(x)$.

## Importance sampling recap

- fundamentally, a trick to help compute expectations in Monte Carlo integration.

- the strategy of sampling from $q$ different from $p$ is
called importance sampling.

- $q$ is the "importance density".
- $w_q(X_i)$ is the "importance weight" of the sampled $X_i$
- the optimal $q(x) \propto h(x)p(x)$ choice is never there – since then
the problem becomes trivial!
- but one tries to choose $q(x)$ such that $\sigma^2_q$ is small

## Posterior mean example

- Suppose we know $p(x) = \frac{1}{c} g(x)$ where the functional form $g$ is
known but the normalizing constant $c$ is not.

- It is easy to see that $c = \int h_c(x)p(x)dx$ where
$h_c(x) = g(x)/p(x)$.

- Therefore, 

$$
\hat{c}_q = \frac{1}{N} \sum_{i = 1}^N w_q(X_i) h_c(X_i) = \frac{1}{N} \sum_{i = 1}^N w_q^*(X_i)
$$

where $w_q^*(x) = g(x) / q(x)$. Notice that 

$$
c = \int g(x) dx = \int w_q^*(x) q(x) dx = \mathbb{E_q}[w_q^*]
$$

. . . 

- Hence a simulation consistent estimate of $V$ is 

$$
\hat{V}_q = \frac{\sum_{i = 1}^N w_q^*(X_i)h(X_i)}{\sum_{i = 1}^N w_q^*}
$$

## Further reading

- A concise review of importance sampling: [Tokdar, Surya T., and Robert E. Kass. "Importance sampling: a review." Wiley Interdisciplinary Reviews: Computational Statistics 2.1 (2010): 54-60.](https://www2.stat.duke.edu/~st118/Publication/impsamp.pdf)

- A natural extension to handle multi-modality: [Neal, Radford M. "Annealed importance sampling." Statistics and computing 11 (2001): 125-139.](https://link.springer.com/article/10.1023/A:1008923215028)

- [Blog post by Radford Neal on "The harmonic mean of the likelihood: worst Monte Carlo method ever"](https://radfordneal.wordpress.com/2008/08/17/the-harmonic-mean-of-the-likelihood-worst-monte-carlo-method-ever/)

- Alternative introduction with some motivation: [Bugallo, Monica F., et al. "Adaptive importance sampling: The past, the present, and the future." IEEE Signal Processing Magazine 34.4 (2017): 60-79.](https://ieeexplore.ieee.org/abstract/document/7974876)

## Acknowledgements

The best parts of these slides were adapted from Prof. Surya Tokdar's notes on Monte Carlo integration.
