---
title: "quiz solutions"
fontsize: 12pt
---

Find here solutions to quizzes with explanations.

<!-- 
::: panel-tabset 
::: 
-->


# quiz 1

::: panel-tabset
### Exercise 1

Is `x` a list or an atomic vector?

```{r}
#| eval: false
x = c(1L, 1)
```

### Solution

`x` is an atomic vector. Specifically, `1L` is coerced to type double due to `c()`.

```{r}
x = c(1L, 1)
typeof(x)
```
:::

::: panel-tabset

### Exercise 2

The output below is the result of printing object `y` to the screen. Is `y` a list or an atomic vector?

```{r}
#| echo: false
f = function(x) {
  return(x ^ 2)
}

list(3, f)
```

### Solution

`y` is a list because it has heterogeneous elements. The first element is the double `3` while the second element is a function.

:::

::: panel-tabset
### Exercise 3
What is the output of the following code chunk:

```{r}
#| eval: false
TRUE | NA
```

### Solution

```{r}
#| echo: false
TRUE | NA
```
:::

::: panel-tabset

### Exercise 4

What's a **generic**?

### Solution

A generic is a function that dispatches to other class-specific functions depending on the class of the argument.
:::

# quiz 2

::: panel-tabset

### Exercise 1

`select()` returns a ___

### Solution

data frame or `tibble`

:::

::: panel-tabset

### Exercise 2

`%>%` and `|>` are both ___

### Solution

pipes
:::

::: panel-tabset

### Exercise 3

`mutate()` creates a new ___

### Solution

column
:::

::: panel-tabset

### Exercise 4

TRUE or FALSE: a tidy data frame has individual observations on the rows, variables on the columns and values in each cell.


### Solution

`TRUE`
:::