---
title: "Lab 1: if it's a function, fix it"
subtitle: "Due Monday January 30 at 5:00pm"  
editor: source
execute:
  eval: false
---

By the end of the lab, you will...

-   write functions and loops with careful checks and control flow

# Getting started

-   If you did not complete [lab 0](/labs/lab0.html), then go back and do so to set up your SSH key, git, and find instructions on how to clone a project.

-   Next in the *terminal* tab, type `cd ~` to navigate to your home directory. Next `git clone git@github.com:sta323-sp23/lab-1-username.git` where `username` is replaced with your username (this is the copied SSH URL from your lab-1 GitHub repo).

-   Navigate to your `lab-1` folder and open the `.Rproj` file.

-   Open the Quarto (`.qmd`) file, change the author name to your name.

## Exercises

For all exercises, you should respond in the space provided in the template `lab-1.qmd` and show all your work. In all answers of type `double`, three significant figures suffices.

### 1.  **Robust quadratic root finding**

The [quadratic formula](https://en.wikipedia.org/wiki/Quadratic_formula) finds $x$ that satisfy the following equation:

$$
ax^2 + bx + c = 0.
$$

In general, $x^*$ such that $f(x^*) = 0$, then $x^*$ is said to be a *root* of $f(x)$. You can find an example of basic implementation of the quadratic formula from [lecture 2: control flow, loops and functions](https://sta323-sp23.github.io/slides/lec02.html#/function-composition).

Write a function named `robustQuadraticRoots()` that accepts three arguments: `a`, `b` and `c` and returns both roots (even if they are repeated) while satisfying the additional cases below. Show the output of your function `robustQuadraticRoots()` on each of the cases below.

```{r}
#| eval: false
# case i 
robustQuadraticRoots(a = 0, b = 3, c = 2)

# case ii
robustQuadraticRoots(a = 0, b = 0, c = 0)

# case iii
robustQuadraticRoots(a = 1, b = 2, c = 3)

# case iv
robustQuadraticRoots(a = 4.3162933554e-11,
                       b = 1.0361118521e6,
                       c = -4.5813932360e5)

# case v
robustQuadraticRoots(a = 0, b = 5, c = 0)
```

::: callout-tip
## Hint
- for case (ii) you should return an error. Since errors will stop your `.qmd` from rendering, you should include in your code chunk `#| error: true`. Read [quarto docs here](https://quarto.org/docs/computations/execution-options.html) for a full list of code chunk execution options.
- for case (iii) you need to deal with complex numbers [read here](https://www.r-tutor.com/r-introduction/basic-data-types/complex)
- for case (iv) the correct answer is $x_1 = -2.4e16$ and $x_2 = -.44$ when rounded to two significant figures.
- for case (v), you should use `warning()` to warn the user that both *a* and *c* are 0.
:::

In each case above, describe why `quadraticRoots()` from the lecture fails.

### 2. **Stirling numbers of the second kind**

Stirling numbers of the second kind count the number of possible partitions of a set of $n$ objects into $k$ disjoint blocks. For example, the number of ways to partition a set of 3 objects in 2 disjoint blocks is $S(3, 2) = 3$ because the set $\{1, 2, 3\}$ can be partitioned into two disjoint blocks in 3 ways:

$$
\begin{aligned}
&\{1, 2\} \cup \{3\},\\
&\{1, 3\} \cup \{2\},\\
&\{2, 3\} \cup \{1\}
\end{aligned}
$$
We can generate Stirling numbers of the second kind, $S(n,k)$, recursively:

$$
S(n,k) = S(n-1, k-1) + kS(n-1, k)
$$

where we start the sequence with $S(n, 1) = 1$ and the boundary condition $S(n, k) = 0$ for $k>n$.

Create a function `ssk()` with arguments `n` and `k` that computes **S**tirling numbers of the **s**econd **k**ind. Print the result of your function for the following inputs:

```{r} 
#| eval: false

# case i
ssk(n = 10, k = 1)

# case ii
ssk(n = 10, k = 10)


# case iii
ssk(n = 100, k = 25)

# case iv
ssk(n = 6, k = 7)
```

::: callout-tip
## Hint
in case (iv) above, your function should print an error. See the hint from the previous exercise for more details.
:::

### 3. **Pareto distribution**

#### Introduction

R provides functions that return useful characteristics of many common 
probability distributions. The naming convention for these functions is a 
prefix, which identifies what the function does, followed by an abbreviation 
of the probability distribution's name. These prefixes are: 

  + `p` for "probability", the cumulative distribution function (CDF)
  + `q` for "quantile", the inverse CDF
  + `d` for "density", the density function (PDF)
  + `r` for "random", to sample a random variable having the specified distribution.

For the normal distribution, these functions are `pnorm`, `qnorm`, `dnorm`, 
and `rnorm`, where the norm portion reminds us this is for the normal 
distribution. For the binomial distribution, these functions are `pbinom`, 
`qbinom`, `dbinom`, and `rbinom`. Click [here](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/Distributions.html)
for a list of probability distributions in the `Base` R package.

The [Pareto distribution](https://en.wikipedia.org/wiki/Pareto_distribution) 
is not available in base R, so we're going to code it ourselves. For this 
lab, we'll just code the quantile function, i.e., `qpareto()`. 
Here's a bit of background on deriving the Pareto's quantile function. 

The Pareto family of distributions is parameterized by $\alpha$ and $x_0$, and 
has probability density function

$$
f(x) = \begin{cases}
\frac{(\alpha - 1)x_0^{\alpha - 1}}{x^{\alpha}}, &x > x_0,\\
0, &x \leq x_0.
\end{cases}
$$

From the PDF it is relatively easy to compute the CDF, which is given by

$$
F(x) = \begin{cases}
0 & x < x_0\\
1 - \left(\frac{x_0}{x} \right)^{\alpha - 1} & x \geq x_0.
\end{cases}
$$

The quantile function is defined for $0 \le p \le 1$, and it returns the 
value $x_p$ such that $F(x_p) = p$. For the Pareto distribution, 
the quantile function is given by 

$$
Q(p) = Q(p, \alpha, x_0) = {x_0}{(1-p)^{-\frac{1}{\alpha - 1}}}.
$$

Using the definition of $Q(p)$, we can compute the $p$th quantile for 
specific values of $p$ manually. For example, to get the median ($p = 0.5$)
of Pareto distributions with $x_0 = 1, \alpha = 3.5$:

```{r}
1 * (1 - 0.5) ^ (-1/(3.5 - 1))
```

It would be helpful to have a function that automated this process,
both so we don't have to remember the form of the quantile function for the 
Pareto distribution, and so we avoid making mistakes.

We will build our function, `qpareto()`, in a sequence of steps.

#### Step 1 `qpareto_1`

Write a function called `qpareto_1()` that takes arguments 
`p`, `alpha`, and `x0` and returns $Q(p, \alpha, x_0)$ as defined above.

::: callout-tip
## Hint
check the result of your function manually (as in the example above) to ensure you have the correct answer.
:::

```{r qpareto_1, eval=FALSE, results='hold'}
qpareto_1(p = 0.5, alpha = 3.5, x0 = 1)
qpareto_1(p = 0.5, alpha = 2.34, x0 = 6e8)
qpareto_1(p = 0.92, alpha = 2.5, x0 = 1e6)
```


#### Step 2

#### `qpareto_2()`

Most of the quantile functions in R have an argument `lower.tail` that is 
either `TRUE` or `FALSE`. If `TRUE`, the function returns the $p$th quantile. 
If `FALSE`, the function returns the $(1-p)$th quantile, i.e., returns the 
value $x_p$ such that $F(x_p) = 1 - p$. 

Create a function `qpareto_2()` that has an additional argument `lower.tail` 
which is by default set to `TRUE`. Your `qpareto_2` function should test
whether `lower.tail` is `FALSE`. If it is `FALSE`, the function should replace
$p$ by $1-p$. Then pass either $p$ or $1-p$ to `qpareto_1()` to compute the
appropriate quantile, i.e., `qpareto_1()` is called from inside of
`qpareto_2()`. Test your function with the two function calls below.

```{r qpareto_2-examples, results='hold'}
#| eval: false
qpareto_2(p = 0.5, alpha = 3.5, x0 = 1)
qpareto_2(p = 0.08, alpha = 2.5, x0 = 1e6, lower.tail = FALSE)
```

There is a downside to writing the function the way we have. 
We need `qpareto_1()`
to be in the work space when `qpareto_2()` is called, 
but there is a big advantage.
If we discover a better way to calculate quantiles of the Pareto
distribution, we
can rewrite `qpareto_1()` and the new version will automatically
be used in `qpareto_2()`.


#### Step 3

#### `qpareto()`

Next, let's add some check with regards to the function's arguments. In the
case of the Pareto quantile function, we need $0\leq p\leq 1$, $\alpha > 1$, 
and $x_0 > 0$.  

Write a function named `qpareto()` that adds these checks to your code from 
function `qpareto_2()`.

Quarto markdown will not compile if your R function stops due to the `stopifnot()`
function. Include `#| error: true` to allow your document to render. See the hint to exercise 1 for more details. 

Test your function on the five function calls below.

```{r qpareto-examples}
#| eval: false
qpareto(p = 0.5, alpha = 3.5, x0 = 1)
qpareto(p = 0.08, alpha = 2.5, x0 = 1e6, lower.tail = FALSE)
qpareto(p = 1.08, alpha = 2.5, x0 = 1e6, lower.tail = FALSE)
qpareto(p = 0.5, alpha = 0.5, x0 = -4)
qpareto(p = 0.5, alpha = 2, x0 = -4)
```

*Is your function vectorized? For which parameters is it vectorized?*

## Style guidelines

All assignments in this course must employ proper coding style, as outlined below:

-   All code should obey the 80 character limit per line (i.e. no code should run off the page when rendering or require scrolling). To enable a vertical line in the RStudio IDE that helps guide this, go to `RStudio` \> `Preferences` \> `Code` \> `Display` \> `Show margin` \> `80`. This will enable a vertical line in your `.qmd` files that shows you where the 80 character cutoff is for code chunks.

-   All commas should be followed by a space.

-   All binary operators should be surrounded by space. For example `x + y` is appropriate. `x+y` is not.

-   All pipes `%>%` or `|>` as well as ggplot layers `+` should be followed by a new line.

-   You should be consistent with stylistic choices, e.g. only use 1 of `=` vs `<-` and `%>%` vs `|>`

-   Your name should be at the top (in the YAML) of each document under "author:"

-   All code chunks should be named (with names that don't have spaces, e.g. `ex-1`, `ex-2` etc.)

-   File names in your GitHub repo such as `lab-x.qmd` must not be changed and left as provided. Additionally, your repo must pass certain basic checks. The results of this check are visible on GitHub via the badges at the top of your README and the actions tab. These are meant to give you feedback around the structure and reproducibility of your repository and assignment - they do not assess the correctness of your work. You should consider them a necessary but not sufficient condition when turning in your work - passing all of the checks simply means your have met a minimum standard of reproducibility for the assignment.

Fundamentally, the check is making sure 1) you only have the files you should in your repository, 2) your `.qmd` renders.

If you have any questions about style, please ask a member of the teaching team.


## Submitting your lab

To submit your assignment, simply commit and push your completed `lab-x.qmd` to your GitHub repo. Your most recent commit 48 hours after the assignment deadline will be graded, and any applicable late penalty will be applied (see [the syllabus](/syllabus.html)). For this reason, **do not push commits** after you are satisfied with your work, or a late penalty will be applied.
